---
title: "Electricity price prediction with machine learning"
output: html_document
date: "2025-04-18"
---

```{r}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# Packages from the link.
library(dplyr)
library(tidyverse)
library(purrr)
library(ggplot2)
# Let's load in our other packages that we usually use.
library(viridis)
library(forcats)
# library(httr)
# library(jsonlite)
library(lubridate)
library(janitor)
library(kableExtra)
library(scales)
library(readxl)
# library(data.table)
# library(plm)
# library(tseries)
# library(stargazer)
library(stats)
# library(Hmisc)
# library(nloptr)
# library(NlcOptim)

# Library to let Rstudio read in python
library(reticulate)
library(fpp2)
library(glmnet)
# library(VGAM)
library(AER)
```


Loading in the raw data below:

```{r}

# Loading in merit data

data <- read_csv("data/student_data.csv")
# Unlike the previous dataset it does not have the column "dispatched"
data <- data |> 
  #data<- merit_aug %>%clean_names%>%
  mutate(dispatched=case_when(dispatched_mw!=0~'Y',.default='N'))


```

I realized that we're not predicting with the recalculated prices that we have. So let's try everything with the prices that we have recalculated.

```{r}

marco_data<-data %>% 
  # filter(date==ymd("2024-04-10"),size>0)%>%
  filter(size>0) %>%
  #select(c('date',"he","actual_posted_pool_price","actual_ail","month",
  #                           "day","hour","year","on_peak","hourly_renewables",
  #                          "temp_yeg","temp_yyc","hourly_imports","hourly_exports",
  #                           "dispatched_mw",'size',"asset_id","dispatched","flexible","nit_settle_cad_gj"))%>%
  group_by(date,he)%>%
  arrange(date,he,actual_posted_pool_price,size,asset_id)%>%
  mutate(block=row_number(),
         merit=cumsum(size),
         mkt_price=max(actual_posted_pool_price*(dispatched=="Y"))
         )%>%
  #fix the estimated ail using the last block in each hour. Assume half the last flexible block is dispatched
  mutate(
    dispatched_mw=case_when(
      # The ~ separates the left and right side of a formula
      # Here, used for the function "case_when()":
        # This line below makes the last dispatched flexible block dispatch half
        block==which(block==max(block*(dispatched=='Y')))*(flexible=='Y') ~ size*.5,
        # Old Code:
        # block==which(block==max(block*(dispatched=="Y")*(flexible=="Y"))) ~ size*.5,
        # block==max(block*(dispatched=="Y")*(flexible=="Y")) ~ size*.5,
        # Here, if it says dispatched then it dispatched full
        dispatched=="Y" ~ size,
        # If it doesn't say dispatched then it's at zero
        dispatched =="N" ~ 0)) %>%
  mutate(ail=sum(dispatched_mw),
  dispatched_merit=cumsum(dispatched_mw))

marco_data <- marco_data |> filter(he!='02*') # Deleting the extra hour from daylight savings for now.



marco_data<-marco_data %>% 
  group_by(date,he)%>%
  mutate(hourly_solar=sum(dispatched_mw*(plant_type=="SOLAR")),
         hourly_wind=sum(dispatched_mw*(plant_type=="WIND")),
          )%>%
  ungroup()

```

Now let's separate our data into the training part, and the testing part.

```{r}

# Making the split 3:1 for now.
#why do you need to do this here? Why
train_data <- marco_data |> filter(year<=2023)
test_data <- marco_data |> filter(year>2023)

```

Now we can start creating a machine learning model. Let's use the methods we learned in class. First, let's try different models and see what fits our regression.

```{r}

# Let's try different models for our prediction model and set them:
#model1 <- as.formula(mkt_price ~ he + month + hourly_renewables + p_natgas + on_peak)
#model2 <- as.formula(mkt_price ~ mktp_daylag)

model3 <- as.formula(mkt_price ~ he + factor(month) + hourly_renewables + nit_settle_cad_gj + on_peak + factor(month)*he + factor(month)*hourly_renewables + factor(month)*on_peak)


# Fit the models
reg1 <- lm(model1, data = marco_data%>%filter(year>=2018,year<=2024))
reg2 <- lm(model2, data = marco_data%>%filter(year>=2018,year<2024))

reg3 <- lm(model3, data = marco_data%>%
             group_by(date,he)|> slice_head(n=1)%>%ungroup()%>%
             select(mkt_price=actual_posted_pool_price,he,month,year,hourly_renewables,nit_settle_cad_gj,on_peak)%>%
             filter(year>=2018,year<2024))

tob1 <- AER::tobit(model3,left=0,right=1000,data = marco_data%>%
             group_by(date,he)|> slice_head(n=1)%>%ungroup()%>%
             select(mkt_price=actual_posted_pool_price,he,month,year,hourly_renewables,nit_settle_cad_gj,on_peak)%>%
             filter(year>=2018,year<2024)
                     )

# model selection 
table1 <- rbind(CV(reg1), CV(reg2), CV(reg3))
rownames(table1) <- c('Model1', 'Model2', 'Model3')

# print nice table
table1 |> 
  kbl(caption = "Model selection using `fpp2::CV()`.") |> 
  kable_styling(bootstrap_options = c("striped", "hover"))

```

Now, let us try to fit a ridge regression.

```{r}

# lambda values
lambdas <- 10^seq(3, -4, length = 100)

# Adding in the temperatures for Edmonton and Calgary causes an error with glmnet().
# That's because glmnet() does not omit these values, so we have to do that ourselves.
train_set <- train_data |> na.omit()
# train_set

# ridge reg for different values of lambda
# We're using the same variables as model 3
est_ridge <- glmnet(
  x = train_set |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix(),
  y = train_set$mkt_price, 
  alpha = 0, 
  standardize = F,
  lambda = lambdas
)


```

Associated with each value of λ is a vector of ridge regression coefficients, stored in a matrix that can be accessed by coef().
 
Now let’s plot the ridge regression coefficients using ggplot2.

```{r}

# organize results
ridge_df <- est_ridge |> coef() |> t() |> as.matrix() |> as.data.frame()
ridge_df <- ridge_df |> dplyr::select(-1) |> mutate(lambda = est_ridge$lambda)
ridge_df <- ridge_df |> gather(key = "variable", value = "coefficient", -lambda)
# plot
ggplot(
  data = ridge_df,
  aes(x = lambda, y = coefficient, color = variable)
) +
geom_line() +
scale_x_continuous(
  expression(lambda),
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
  trans = "log10"
) +
scale_y_continuous("Ridge coefficient") +
scale_color_viridis_d("Predictor", option = "magma", end = 0.8) +
theme_minimal() +
theme(legend.position = "bottom")

```

We now use cross-validation to choose the tuning parameter $\lambda$. Choosing the number of folds to be 10, since we have 10 years in our training data.

```{r}

# ridge cv
ridge_cv <- cv.glmnet(
  x = train_set |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix(),
  y = train_set$mkt_price, 
  alpha = 0,
  standardize = F,
  lambda = lambdas,
  type.measure = "mse",
  nfolds = 10
)

```

And now we plot to see the the appropriate $\lambda$.

```{r}

# create data frame with results
ridge_cv_df <- data.frame(
  lambda = ridge_cv$lambda, 
  rmse = sqrt(ridge_cv$cvm)
)
# plot
ggplot(
  data = ridge_cv_df,
  aes(x = lambda, y = rmse)
) +
geom_line() +
geom_point(
  data = ridge_cv_df |> filter(rmse == min(rmse)),
  size = 3.5,
  color = "darkorange"
) +
scale_y_continuous("RMSE") +
scale_x_continuous(
  expression(lambda),
  trans = "log10",
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
) +
theme_minimal()

```

Now let's run the ridge regression, with the minimum value for Lambda, and store those values.

```{r}

# Let's fit the ridge regression on the test data.
final_ridge <- glmnet(
  x = train_set |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix(),
  y = train_set$mkt_price, 
  alpha = 0,
  standardize = F,
  lambda = ridge_cv$lambda.min
)

# predict with ridge regression
pred_ridge <- predict(
  final_ridge,
  type = "response",
  s = ridge_cv$lambda.min,
  newx = test_data |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix()
)

```

Now let's check on a lasso regression to see if it will yield even better predictions.

```{r}

# lasso reg for different values of lambda
est_lasso <- glmnet(
  x = train_set |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix(),
  y = train_set$mkt_price,
  alpha = 1, 
  standardize = F,
  lambda = lambdas
)

```

Associated with each value of $\lambda$ is a vector of lasso regression coefficients, stored in a matrix that can be accessed by coef().

Now let’s plot the lasso regression coefficients using ggplot2.

```{r}

# organize results
lasso_df <- est_lasso |> coef() |> t() |> as.matrix() |> as.data.frame()
lasso_df <- lasso_df |> dplyr::select(-1) |> mutate(lambda = est_lasso$lambda)
lasso_df <- lasso_df |> gather(key = "variable", value = "coefficient", -lambda)
# plot
ggplot(
  data = lasso_df,
  aes(x = lambda, y = coefficient, color = variable)
) +
geom_line() +
scale_x_continuous(
  expression(lambda),
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
  trans = "log10"
) +
scale_y_continuous("Lasso coefficient") +
scale_color_viridis_d("Predictor", option = "magma", end = 0.8) +
theme_minimal() +
theme(legend.position = "bottom")

```

Cross validation to tune the $\lambda$ parameter. We again use the built-in cross-validation function cv.glmnet().

```{r}

# lasso cv
lasso_cv <- cv.glmnet(
  x = train_set |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix(),
  y = train_set$mkt_price, 
  alpha = 1,
  standardize = F,
  lambda = lambdas,
  type.measure = "mse",
  nfolds = 10
)

```

From our cross-validation, we graph the RMSE of the lambdas.

```{r}

# create data frame with results
lasso_cv_df <- data.frame(
  lambda = lasso_cv$lambda, 
  rmse = sqrt(lasso_cv$cvm)
)
# plot
ggplot(
  data = lasso_cv_df,
  aes(x = lambda, y = rmse)
) +
geom_line() +
geom_point(
  data = lasso_cv_df |> filter(rmse == min(rmse)),
  size = 3.5,
  color = "darkorange"
) +
scale_y_continuous("RMSE") +
scale_x_continuous(
  expression(lambda),
  trans = "log10",
  labels = c("0.1", "10", "1,000", "100,000"),
  breaks = c(0.1, 10, 1000, 100000),
) +
theme_minimal()

```

Now let's predict with the lasso regression and see how it does.

```{r}

# fit lasso regression 
final_lasso <- glmnet(
  x = train_set |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix(),
  y = train_set$mkt_price, 
  alpha = 1,
  standardize = F,
  lambda = lasso_cv$lambda.min
)

# predict with lasso regression
pred_lasso <- predict(
  final_lasso,
  type = "response",
  s = lasso_cv$lambda.min,
  newx = test_data |> dplyr::select(-c('mkt_price','date','hour','day','on_peak','year',"ail","temp_yeg","temp_yyc","hourly_imports","hourly_exports")) |> as.matrix()
)

```

Let's look at the mean squared errors (MSE), and then we can decide on a model to use.

```{r}

test_data<-marco_data%>%
             group_by(date,he)|> slice_head(n=1)%>%ungroup()%>%
             select(mkt_price=actual_posted_pool_price,he,month,year,hourly_renewables,nit_settle_cad_gj,on_peak)%>%
             filter(year==2024)


# Make an object that will store the errors
cv.error <- rep(NA,7)

# Predict with our first model:

# 1) First we predict with our tobit model and store them:
mu <- predict(tob1,newdata=test_data)
sigma <- tob1$scale
# 2)
p0 <- pnorm(mu/sigma)
# 3)
lambda <- function(x) dnorm(x)/pnorm(x)
ey0 <- mu + sigma * lambda(mu/sigma)
# 4) This last step gives us the actual expected prices
ey <- p0 * ey0
# So we just equate again:
pred_tob <- ey

# And now we add our tobit regression predicted values to check the MSE:
cv.error[7] <- mean((test_data$mkt_price - pred_tob)^2)

print(cv.error, digits = 4)

```

Let's now try getting the predictions and see how our model did. It looks like model 3 is the best at first, based on the AIC and BIC, however our ridge regression might be better.

```{r}

# Let's create a new dataframe to make graphing easier:
for_graph <- test_data |> select(date,he,mkt_price,mktp_daylag)

# Then let's get our predictions:

pred1 <- predict(reg1, newdata = test_data)
for_graph <- for_graph |> mutate(pred1 = pred1)
pred2 <- predict(reg2, newdata = test_data)
for_graph <- for_graph |> mutate(pred2 = pred2)
pred3 <- predict(reg3, newdata = test_data)
for_graph <- for_graph |> mutate(pred3 = pred3)
# For the tobit predictions, we already did them in code above, so we're just storing here.
for_graph <- for_graph |> mutate(pred_tob=pred_tob)
# And now we put in our predictions from the ridge regression:
for_graph <- for_graph |> mutate(pred_ridge=pred_ridge[,1])
# And now from the lasso regression:
for_graph <- for_graph |> mutate(pred_lasso=pred_lasso[,1])

# Let's save this as a csv and graph on python:
for_graph |> write_csv("graphing_predictions.csv")

```


Here I'm making code to create in-sample estimations with the tobit model.

```{r}

# TOBIT PREDICTIONS CALCULATED MANUALLY:
# Like above, we are calculating the estimated values.
# https://stats.stackexchange.com/questions/149091/censored-regression-in-r

# 1) First we predict with our tobit model and store them:
mu <- predict(tob1,newdata=test_data)
sigma <- tob1$scale
# 2)
p0 <- pnorm(mu/sigma)
# 3)
lambda <- function(x) dnorm(x)/pnorm(x)
ey0 <- mu + sigma * lambda(mu/sigma)
# 4) This last step gives us the actual expected prices
ey <- p0 * ey0
# So we just equate again:
pred_tob_insample <- as_tibble(ey)

test_data<-test_data %>% bind_cols(pred_tob_insample)

ggplot(test_data)%>%
  geom_line()


# And now let's create a dataset where we'll compare with in sample values.
price_w_tob_insample <- train_data |> select(date,he,mkt_price,mktp_daylag) |>
  mutate(pred_tob=pred_tob_insample)

# Now let's save this in a CSV:
price_w_tob_insample |> write_csv("insample_predictions.csv")


```

Now that we've chosen a model, let's see if we need to retrain our model in a method of a rolling window.

```{r}

# If we will be looping on our dates and variables, it would be useful to have a datetime variable.
# Let's create one:

# Creating a new date and time variable just in case. It might be important.
fulldata <- fulldata |> mutate(datetime = format(make_datetime(year = year, hour = round(as.numeric(hour)), day = day, month = month), "%Y-%m-%d %H:%M:%S"))

# Let's create a list of all the unique dates in our dataset.
date_list <- fulldata |> select(date) |> unique()
# And now we can use this to filter things out.
year(date_list$date[length(date_list$date)]) # Just as an example of what we can do.

# Now we need to create a loop that will loop and change the training window, then change the test day, as it goes along.
# Let's get the length of time (in days) for how long our training set is supposed to be:
train_length <- train_data |> select(date) |> unique() |> nrow()
test_length <- test_data |> select(date) |> unique() |> nrow()

date_list$date[train_length]

# Let's make an empty list for the tobit predictions:
pred_tob_roll <- list()

# Now we loop and predict on a rolling window that updates:
for (i in 1:test_length){
  # Filter out the relevant training data
  train_temp <- fulldata |> filter(date<=date_list$date[train_length-1+i]) # |> filter(date>=date_list$date[i])
  # Filter out the day that we are forecasting
  test_temp <- test_data |> filter(date==date_list$date[train_length+i])
  
  # Now we train our tobit regression and fit it:
  tob_loop <- AER::tobit(model3,left=0,right=1000,data=train_temp)
  # Then we run it on our forecast data. Recall, for Tobit, we have to manually calculate the predicted values each time:
  #----------------------------------------------------------
  # 1) First we predict with our tobit model and store them:
  mu <- predict(tob_loop,newdata=test_temp)
  sigma <- tob_loop$scale
  # 2)
  p0 <- pnorm(mu/sigma)
  # 3)
  lambda <- function(x) dnorm(x)/pnorm(x)
  ey0 <- mu + sigma * lambda(mu/sigma)
  # 4) This last step gives us the actual expected prices
  ey <- p0 * ey0
  # So we just equate again:
  pred_tob <- ey
  #----------------------------------------------------------
  # And now we need to store our 24 hour prediction:
  pred_tob_roll <- append(pred_tob_roll,pred_tob)
}

price_table <- test_data |> select(c("date","he","mkt_price")) |>
  mutate(pred_tob_roll=as.numeric(pred_tob_roll))
for_graph <- for_graph |> mutate(pred_tob_roll=as.numeric(pred_tob_roll))

# Now let's check how accurate this is:

mean((for_graph$mkt_price - for_graph$pred_tob_roll)^2)

```









